Endogenous Politics and \\ the Design of Trade Institutions

### 2018-03-01 presentation.tex
significant revisions to presentation for Nebraska / new pitch

\usepackage{appendixnumberbeamer}

\title[Endogenous Politics and the Design of Trade Institutions\hspace{2.05in}\insertframenumber/\inserttotalframenumber]{Endogenous Politics and \\ the Design of Trade Institutions}
%\author[Kristy Buzard]{Kristy Buzard \\ Syracuse University and The Wallis Institute \\ kbuzard@syr.edu}
\author[Kristy Buzard]{\texorpdfstring{Kristy Buzard\newline Syracuse University \newline\url{kbuzard@syr.edu}}{Kristy Buzard}}
\date{January 6, 2017}
\begin{document}
\maketitle
%\insertpresentationendpage removed b/c of appendix

@@

\pause
\begin{enumerate}[<+->]
\item When is endogenizing political pressure important for answering optimal design questions?
	\begin{itemize}
		\item Exogenous vs. endogenous politics
	\end{itemize}
	\vskip.1in
\item Can trade agreements be used to manipulate domestic lobbying incentives?
	\begin{itemize}
		\item Government objective function
	\end{itemize}
\end{enumerate}

@@

\begin{frame}{Preview}
\frametitle{Political Economy of Trade Institutions}
\pause
With a few exceptions, TA design literature has taken political economy forces to be exogenous. I:
\pause
\begin{itemize}[<+->]
	\item endogenize politics into a standard model for studying TA design questions
	\item carefully distinguish between dynamics induced by exogenous and endogenous politics for
		\begin{itemize}[<+->]
			\item base case with tariff caps
			\item tariff caps with escape clause
		\end{itemize}
	\item examine escape clause design when both exogenous and endogenous forces are present
\end{itemize}
\end{frame}

@@

\pause
\begin{itemize}[<+->]
	\item Show that TAs may be used to manipulate domestic political actors (no long-run distortions)
	\item For both tariff caps and escape clauses, outcomes are very different with endogenous politics
	\item Demonstrate that (standard, theoretical) escape clause can't work in the presence of endogenous political pressure
		\begin{itemize}
			\item Points to real-world design of WTO Agreement on Safeguards
			\item May explain why escape clause has fallen out of use
		\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\includegraphics[height=2.75in, width=4.25in]{weight.jpg}
\end{frame}

@@

{\begin{frame}{Tariff Caps: Exogenous vs. Endogenous $\ga$}
\pause
Must set tariff at or below specified level (aka tariff cap)
\pause
\begin{itemize}[<+->]
	\item $\ga$ exogenous (Bagwell $\&$ Staiger 2005): Negotiated weak bindings (a) are higher than those gov'ts would choose if they instead negotiated strong bindings and (b) imply that governments with low realizations of $\ga$ set their applied tariffs strictly below the bound level.

	\item $\ga$ endogenous: Governments will not set applied tariffs strictly below the bound level. They may use the weak tariff binding either to encourage and/or restrain endogenous political pressure.

@@

\begin{frame}[label=self]
\frametitle{Tariff Caps with Self Enforcement}
\pause
\begin{itemize}[<+->]
	\item $\ga$ exogenous (Bagwell $\&$ Staiger 2005): if governments patient enough ($\de$ high enough), optimal externally-enforced weak binding can be self-enforced
  \item $\ga$ endogenous: optimal externally-enforced weak binding may not be self-enforcing
		\begin{itemize}
			\item Problem: lobby is an additional repeated-game player
			\item Lobby's incentive constraint is harder to satisfy as $\de$ increases
		\end{itemize}

\end{itemize}
\pause
\hyperlink{repeated<1>}{\beamergotobutton{Repeated Game Intuition}}
\end{frame}

@@

@@

\begin{frame}{Escape Clause with Endogenous Politics}
When $\ga$ is \textit{only} endogenous:
\pause
\begin{itemize}[<+->]
	\item Benefit of escape clause from exogenous case is gone
	\item Assuming lower binding is set to maximize political welfare, escape clause encourages inefficiently high lobbying effort / protection
	\item (Incentive compatibility still an issue, but often not the central one)
		\begin{itemize}[<+->]
			\item (If lobby's preferred tariff $\geq$ escape clause binding, gov't experiences high $\ga$, no need to lie)
		\end{itemize}
  \end{itemize}

@@

\begin{frame}{Conclusion}
Taking into account endogenous political forces alongside exogenous ones...
\pause
\begin{itemize}[<+->]
		\item helps explain the structure and enforcement of the WTO Safeguards measure
		\item can help us think about optimal design of trading insitutions
		\item demonstrates that TAs can be used to discourage lobbing activity in general
		\item provides additional general explanation for tariff caps
\end{itemize}

@@

\appendix
\begin{frame}[label=repeated]
\frametitle{Repeated Game Intuition}
\pause
Legislature: break agreement if punishment not strong enough
\pause
\begin{itemize}
	\item i.e. if one period of gain from cheater's payoff is greater than T-periods of loss from trade-war
\end{itemize}

\pause
\vskip.1in
Executives: set lowest $\bta$ that makes paying $\ov{e}_b$ unprofitable \textit{and} satisfies legislature's condition
\pause
\begin{itemize}
	\item[$\Rightarrow$] $e_b=0$, agreement remains in force
	\pause
	\item High tariffs, no lobbying, no trade disruptions
\end{itemize}
\hyperlink{self<1>}{\beamergotobutton{Go Back}}
\end{frame}

>>>>

%\usepackage{appendixnumberbeamer}

\title[Endogenous Politics and the Design of Trade Institutions\hspace{2.05in}\insertframenumber/\inserttotalframenumber]{Endogenous Politics and \\ the Design of Trade Institutions}
%\author[Kristy Buzard]{Kristy Buzard \\ Syracuse University and The Wallis Institute \\ kbuzard@syr.edu}
\author[Kristy Buzard]{\texorpdfstring{Kristy Buzard\newline Syracuse University \newline\url{kbuzard@syr.edu}}{Kristy Buzard}}
\date{March 2, 2018}
\begin{document}
\maketitle
%\insertpresentationendpage removed b/c of appendix

@@

\pause
\begin{enumerate}[<+->]
\item Can trade agreements be used to manipulate domestic lobbying incentives?
	\begin{itemize}
		\item Government objective function
	\end{itemize}
\vskip.1in
	\item When is endogenizing political pressure important for answering optimal design questions?
	\begin{itemize}
		\item Exogenous vs. endogenous politics
	\end{itemize}
\end{enumerate}

@@

\begin{frame}{Preview}
\frametitle{Political Economy of Trade Institutions}
\pause
With a few exceptions, TA design literature has taken political economy forces to be exogenous. I:
\pause
\begin{itemize}[<+->]
	\item endogenize politics into a standard model for studying TA design questions
		\begin{itemize}
			\item use this to examine gov't objective
		\end{itemize}
	\item carefully distinguish between dynamics induced by exogenous and endogenous politics for
		\begin{itemize}[<+->]
			\item base case with tariff caps
			\item tariff caps with escape clause
		\end{itemize}
	\item examine escape clause design when both exogenous and endogenous forces are present
\end{itemize}
\end{frame}

@@

\pause
\begin{itemize}[<+->]
	\item TAs may be used to manipulate domestic political actors (no long-run distortions)
	\item Simple modeling framework can capture results from models of both exogenous and endogenous politics
		\begin{itemize}
			\item For both tariff caps and escape clauses, endogenous politics changes outcomes dramatically
		\end{itemize}
	\item Standard, theoretical escape clause can't work in the presence of endogenous political pressure
		\begin{itemize}
			\item Points to real-world design of WTO Agreement on Safeguards
			\item May explain why escape clause has fallen out of use
		\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\includegraphics[height=2.75in, width=4.25in]{weight.jpeg}
\end{frame}

\begin{frame}{Comparison to the Literature}
Others derive non-monotonicity in lobbying effort/tariff 
\begin{itemize}
	\item Maggi $\&$ Rodriguez-Clare (1998/2007): dynamic firm investment disortions
	\item Mitra (2002): lobbies pay investment cost to form
\end{itemize}

\vskip.15in
Here, I acknowledge the gov't objective function may be fundamentally non-monotonic
\begin{itemize}
	\item Achieve same results with simpler model
	\item Endogenous politics in a wider range of questions
	\item Can have both endogenous / exogenous at the same time $\Rightarrow$ unify the exogenous and endogenous politics literatures
\end{itemize}
\end{frame}

@@

{\begin{frame}{Tariff Caps: Exogenous vs. Endogenous $\ga$}
\pause
Must set tariff at or below specified level (aka tariff cap)
\pause
\begin{itemize}[<+->]
	\item $\ga(s)$, i.e. exogenous: Negotiated weak bindings (a) are higher than those gov'ts would choose if they instead negotiated strong bindings and (b) imply that governments with low realizations of $\ga$ set their applied tariffs strictly below the bound level.
		\begin{itemize}
			\item Bagwell $\&$ Staiger 2005 result
		\end{itemize}

	\item $\ga(e)$, i.e. endogenous: Governments will not set applied tariffs strictly below the bound level. They may use the weak tariff binding either to encourage and/or restrain endogenous political pressure.
		\begin{itemize}
			\item Maggi $\&$ Rodriguez-Clare (1998/2007) result
		\end{itemize}

@@

\begin{frame}[label=self]
\frametitle{Tariff Caps with Self Enforcement}
\pause
\begin{itemize}[<+->]
	\item $\ga$ exogenous (Bagwell $\&$ Staiger 2005): if governments patient enough (discount factor $\de$ high enough), optimal externally-enforced weak binding can be self-enforced
  \item $\ga$ endogenous: optimal externally-enforced weak binding may not be self-enforcing
		\begin{itemize}
			\item Problem: lobby is an additional repeated-game player
			\item Lobby's incentive constraint is harder to satisfy as $\de$ increases
		\end{itemize}

\end{itemize}
\pause
%\hyperlink{repeated<1>}{\beamergotobutton{Repeated Game Intuition}}
\end{frame}

@@

\begin{frame}[label=repeated]
\frametitle{Repeated Game Intuition}
\pause
Legislature: break agreement if punishment not strong enough
\pause
\begin{itemize}
	\item i.e. if one period of gain from cheater's payoff is greater than T-periods of loss from trade-war
\end{itemize}

\pause
\vskip.1in
Executives: set lowest $\bta$ that makes paying $\ov{e}_b$ unprofitable \textit{and} satisfies legislature's condition
\pause
\begin{itemize}
	\item[$\Rightarrow$] $e_b=0$, agreement remains in force
	\pause
	\item High tariffs, no lobbying, no trade disruptions
\end{itemize}
%\hyperlink{self<1>}{\beamergotobutton{Go Back}}
\end{frame}

\begin{frame}{When the world is more complicated...}
Now suppose political pressure is a result of both endogenous and exogenous forces (i.e. $\ga(s,e)$):
\pause
\vskip.1in
\begin{beamerboxesrounded}[upper=palette tertiary, shadow=true]{Water in the Bindings with $\ga(s,e)$}
  Assume $\ga(s,e) = \ga(s) + \ga(e)$. In order for governments to set applied tariffs strictly below the weak binding, the low shock must be realized \textit{and} the lobby must not have the incentive to `top up' political pressure to the level associated with the binding. 
\end{beamerboxesrounded}

\pause
\begin{itemize}[<+->]
	\item Can happen if gov't mis-judges lobby's incentives
	\item In general, gov't prefers cap because lobby will `fill in' for low shock up to gov'ts optimal level of $\ga$
\end{itemize}
\end{frame}

@@

\begin{frame}{Escape Clause with Endogenous Politics}
When $\ga$ is \textit{only} endogenous:
\pause
\begin{itemize}[<+->]
	\item Benefit of escape clause from exogenous case is gone
	\item Assuming lower binding is set to maximize political welfare, escape clause encourages inefficiently high lobbying effort / protection
	\item Incentive compatibility still an issue, but often not the central one
		\begin{itemize}[<+->]
			\item If lobby's preferred tariff $\geq$ escape clause binding, gov't experiences high $\ga$, no need to lie
		\end{itemize}
  \end{itemize}

@@

\begin{frame}{Conclusion}
Taking into account endogenous political forces alongside exogenous ones in this simplified modeling framework
\pause
\begin{itemize}[<+->]
		\item demonstrates that TAs can be used to discourage lobbing activity in general
		\item can nest established results and provide new insights
		\item can answer questions about optimal design of trading institutions more fully
			\begin{itemize}
				\item provides additional general explanation for tariff caps
				\item helps explain the structure and enforcement of the WTO Safeguards measure
			\end{itemize}
\end{itemize}

@@
